Theory of surface second-harmonic generation for semiconductors including effects of nonlocal operators

### 2014-12-03 manuscript.tex

\documentclass[floatfix,prb,aps,superscriptaddress,showpacs,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{ulem}
\usepackage{subfigure}
\usepackage{bm}
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%\usepackage{showlabels}
%\usepackage{showkeys}
\allowdisplaybreaks[1]%biutiful equation breaker!!!

%%%
% notes:
% check for tense consistency
% Local fields?
%%%

@@

\begin{abstract}
We formulate a theoretical approach of surface second-harmonic generation 
from semiconductor surfaces based on the length gauge. 
Within the independent particle approximation 
the nonlinear second-order surface susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$   
is calculated, including for the first time in one unique formulation
i) the scissors correction, needed to have the correct value of the
energy band gap,
ii) the derivation for the inclusion of the cut function, 
used to extract the surface nonlinear response,
and iii) the contribution of the nonlocal part of the 
pseudopotentials, routinely  used in \textit{ab initio} band structure calculations. 
First and third terms are described by spatially nonlocal quantum
mechanical operators and are fully taken into account in the present formulation. 
As a test-case of the approach we calculate 
$\chi^{xxx}(-2\omega;\omega,\omega)$ for the clean Si(001)$2\times 1$
reconstructed surface and find that our scheme to extract the surface
through the cut function
readily works and is exact. The effects of the scissors correction and
of the nonlocal part of the pseudopotentials are discussed for the
first time in surface nonlinear optics. 
The scissors correction shifts the spectrum to 
higher energies though the shifting is not rigid and mixes the 
$1\omega$ and $2\omega$ resonances, 
and has a strong influence in the line-shape. 
The 
effects of the nonlocal part of the pseudopotentials 
keeps the same line-shape of $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$, but 
reduces its value
by 15-20\%.  
Therefore, the inclusion of the three aforementioned 
contributions is very important and 
makes our scheme unprecedented and opens the possibility to 
study SSHG with more versatility and providing more accurate results 
than previous approaches. 
\end{abstract}

% \begin{abstract}
% We formulate a theoretical approach of surface second-harmonic generation from 
% semiconductor surfaces based on the length gauge. We calculate the nonlinear 
% second-order surface susceptibility tensor 
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
% independent particle approximation, and for the first time in one unique 
% formulation we include i) the scissors correction, needed to have the correct 
% value for the energy band gap, ii) the derivation for the inclusion of the cut 
% function used to extract the surface nonlinear response, and iii) the 
% contribution of the nonlocal part of the pseudopotentials that are routinely 
% used in \textit{ab initio} band structure calculations. The first and third 
% terms are described by spatially nonlocal quantum mechanical operators and 
% are fully taken into account in the present formulation. As a test-case for 
% this approach, we calculate $\chi^{xxx}(-2\omega;\omega,\omega)$ for a clean 
% Si(001)$2\times 1$ reconstructed surface and find that this scheme for 
% extracting the surface response through the cut function is exact and readily
% implemented. The effects of the scissors correction and the nonlocal part of 
% the pseudopotentials are discussed for the first time for surface nonlinear 
% optics. The scissors correction non-rigidly shifts the spectrum to higher 
% energies by mixing the $1\omega$ and $2\omega$ resonances and exerts a strong 
% influence on the line-shape. The effects of the nonlocal part of the 
% pseudopotentials keeps the same line-shape of 
% $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$ but reduces its value by 
% 15-20\%. Therefore, our scheme is unprecedented through the inclusion of 
% these three contributions and also opens the possibility to study SSHG with 
% more versatility and more accuracy than previous approaches. 
% \end{abstract}

\maketitle
%\tableofcontents

@@

In recent years surface nonlinear optical spectroscopies, particulary surface 
second-harmonic generation (SSHG), have evolved as useful nondestructive 
and noninvasive tools to study surface and interface properties. These 
properties include atomic structure, phase transitions, adsorption of 
atoms, and many others.\cite{daumPRL93, mcgilpOE94, meyerPRL95, powerPRL95, 
godefroyAPL96, hoferAPA96, dadapPRB97, bloembergenAPB99, mcgilpSRL99, 
suzukiAPB99, mitchellSS01, hughesPRB96, guyotPRB88, downerPSSA01, shenAPB99, 
shenNAT89, chenPRL81, mendozaPRL98, downerSIA01} Nowadays, SSHG spectroscopy 
is a crucial tool for research and development in microelectronics,
\cite{zheltikovLP00} semiconductors, \cite{lupkeSSR99} nanomaterials,
\cite{salazar-aparicioPRB14} and many more recent areas of great scientific 
and commercial interest.\cite{cazzanelliNM14} The high surface sensitivity 
of SSHG spectroscopy is due to the fact that within the dipole approximation 
the bulk SHG signal of centrosymmetric materials is identically zero. 
Radiation can occur only at the surface where the inversion symmetry is 
broken. Much of the foundation of surface science has been built from 
experiments involving emission or scattering of electrons from surfaces. 
These require ultrahigh vacuum (UHV) environments and provide no access to 
buried interfaces. However, SSHG is compatibile with non-UHV conditions and 
has access to interfaces buried beneath transparent overlayers. Even when 
applied to surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

The usefulness of SSHG could be limited by the lack of microscopic theoretical
understanding of the nonlinear spectra. The macroscopic phenomenological 
theory of SSHG, which relates the intensity, phase, and polarization of 
detected fields to nonlinear and linear susceptibilities at the material 
interface is now fully developed. However, microscopic theory that relates 
atomic-level structure to the nonlinear source polarization is still being 
developed. In this article we put forward an approach to calculate the
microscopic second-harmonic surface susceptibility that encompasses several
theoretical features not taken into account before. Within the independent 
particle approximation (IPA) some frameworks for bulk SHG have been developed 
to study the nonlinear optical response of bulk materials with varying degrees 
of sophistication.\cite{butcherPOPS63, aspnesPRB72, sipePRB93, levinePRB94,
aversaPRB95, hughesPRB96, rashkeevPRB98}

The most used framework for \textit{ab initio} calculations, Density 
Functional Theory (DFT) within the Local Density Approximation (LDA),
\cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
well understood that one has to include the many-body interaction to correct 
for this underestimation of the gap. In this context, the so-called GW 
approximation\cite{onidaRMP02} is known to correct the band gap of most 
semiconductors\cite{luceroJPCM12}. However this can be a very expensive
calculation and thus one uses the much simpler scissors operator scheme.
\cite{levinePRL89,levinePRL91,delsolePRB93}  
This allows us to ``open'' the DFT-LDA gap to 
its correct experimental or GW value even when experiments are not available. 
In this context, to correct for the underestimation of the energy band gap of 
semiconductors Nastos et al.\cite{nastosPRB05} used the ``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to correctly include the 
many-body corrections through the scissors operator in the SH susceptibility.
Later, Cabellos \textit{et al}.\cite{cabellosPRB09} elaborated a derivation 
of the ``velocity gauge'' or ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the 
scissors operator and proved gauge invariance with respect to the length 
gauge. From these works it is clear the length gauge is a much better staring
point to obtain the surface SH (SSH) susceptibility, as will be elaborated 
in this article. However, these considerations are only valid for bulk 
semiconductors.

Concerning the optical response of surfaces and interfaces, Reining 
\textit{et al.}\cite{reiningPRB94} introduced the concept of a cut function 
in order to obtain the surface SH susceptibility tensor. This cut function 
is required since one usually uses a slab approach when treating semi-infinite 
surface systems. If the slab is centrosymmetric the susceptibility tensor will 
be identically zero. The cut function is such that it separates the nonlinear 
response for the two surfaces of the slab avoiding the destructive 
interference between them giving a finite value that one identifies with the 
SSH susceptibility tensor. If the slab is not centrosymmetric the cut function 
can be used to separate the different signals coming from either surface of 
the slab. Indeed, one of the main results of this article is to show that the 
SSH  susceptibility tensor obtained by using the cut function is correctly 
extracted from the slab. After Reining \textit{et al.},\cite{reiningPRB94}
Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02,sanoPRB02}
followed 
upon this work and in particular Ref.~\onlinecite{arzatePRB01} went into a 
detailed analysis of the different contributions to the SHG spectra of a 
surface and the nuanced relationship between bulk, surface, interband, 
intraband, $1\omega$ and $2\omega$ terms, and
Ref. \onlinecite{mejiaRMF04} developed a layer-by-layer 
analysis for the nonlinear responses of semiconductor systems, within a 
tight-binding framework. This model allows for obtaining results from 
selected regions of a system including the surface. However, in these 
references the scissors operator is either excluded or incorrectly 
implemented.

Finally, DFT-LDA calculations are often based on the use of pseudopotentials.
As it will be discussed 
in this article, the presence of a nonlocal part of the
pseudopotential introduces
corrections to
the momentum operator of the electron that have to be included with care in 
the SSH susceptibility. So far, this contribution has been neglected in the 
literature.

Therefore, within the IPA the most complete approach for the calculation of 
the SSH susceptibility is one which includes i) the scissors correction, 
ii) the cut function, and iii) the contribution of the nonlocal part of the 
pseudopotential. Therefore, the goal of this paper is to derive a new 
expression within the length gauge for the SSH susceptibility tensor 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ 
that include the aforementioned contributions. The inclusion of these three 
contributions makes our scheme unprecedented and opens the possibility to 
study surface SHG with more versatility and providing more accurate results 
than previous approaches.

The paper is organized as follows. In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
length gauge formalism. This derivation includes the addition of the terms 
mentioned above that have been absent in previous works. Also, our 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ can be used 
for a layer-by-layer analysis if desired. In Sec. \ref{results} we show the 
results of this reformulation with a study of a clean Si(001) surface with a 
$2 \times 1$ reconstruction, 
proving the correctness of our approach with a special test-case,
never exploited before.
We compare results from before and after adding 
the different nonlocal contributions. Finally, in Sec. \ref{conc} we give 
our conclusions.

% \section{Introduction}\label{intro}

% In recent years surface nonlinear optical spectroscopies, particulary surface 
% second-harmonic generation (SSHG), have evolved as useful nondestructive 
% and noninvasive tools to study surface and interface properties. These 
% properties include atomic structure, phase transitions, adsorption of 
% atoms, and many others.\cite{daumPRL93, mcgilpOE94, meyerPRL95, powerPRL95, 
% godefroyAPL96, hoferAPA96, dadapPRB97, bloembergenAPB99, mcgilpSRL99, 
% suzukiAPB99, mitchellSS01, hughesPRB96, guyotPRB88, downerPSSA01, shenAPB99, 
% shenNAT89, chenPRL81, mendozaPRL98, downerSIA01} Nowadays, SSHG spectroscopy 
% is a crucial tool for research and development in microelectronics,
% \cite{zheltikovLP00} semiconductors, \cite{lupkeSSR99} nanomaterials,
% \cite{salazar-aparicioPRB14} and many more recent areas of great scientific 
% and commercial interest.\cite{cazzanelliNM14} The high surface sensitivity 
% of SSHG spectroscopy is due to the fact that within the dipole approximation 
% the bulk SHG signal of centrosymmetric materials is identically zero. 
% Radiation can occur only at the surface where the inversion symmetry is 
% broken. Much of the foundation of surface science has been built from 
% experiments involving emission or scattering of electrons from surfaces. 
% These require ultrahigh vacuum (UHV) environments and provide no access to 
% buried interfaces. However, SSHG is compatibile with non-UHV conditions and 
% has access to interfaces buried beneath transparent overlayers. Even when 
% applied to surfaces in UHV, the light source and detectors can be aligned and
% used outside the vacuum chamber.

% The usefulness of SSHG could be limited by the lack of microscopic theoretical
% understanding of the nonlinear spectra. The macroscopic phenomenological 
% theory of SSHG, which relates the intensity, phase, and polarization of 
% detected fields to nonlinear and linear susceptibilities at the material 
% interface is now fully developed. However, microscopic theory that relates 
% atomic-level structure to the nonlinear source polarization is still being 
% developed. In this article we put forward an approach to calculate the
% microscopic second-harmonic surface susceptibility that encompasses several
% theoretical features not taken into account before. Within the independent 
% particle approximation (IPA) some frameworks for bulk SHG have been developed 
% to study the nonlinear optical response of bulk materials with varying degrees 
% of sophistication.\cite{butcherPOPS63, aspnesPRB72, sipePRB93, levinePRB94,
% aversaPRB95, hughesPRB96, rashkeevPRB98}

% The most used framework for \textit{ab initio} calculations, Density 
% Functional Theory (DFT) within the Local Density Approximation (LDA),
% \cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
% well understood that one has to include the many-body interaction to correct 
% for this underestimation of the gap. In this context, the so-called GW 
% approximation\cite{onidaRMP02} is known to correct the band gap of most 
% semiconductors\cite{luceroJPCM12}. However this can be a very expensive
% calculation and thus one uses the much simpler scissors operator scheme.
% \cite{levinePRL89,levinePRL91,delsolePRB93} This allows us to ``open'' the 
% DFT-LDA gap to its correct experimental or GW value even when experiments 
% are not available. In this context, to correct for the underestimation of 
% the energy band gap of semiconductors Nastos et al.\cite{nastosPRB05} used 
% the ``length gauge'' or ``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to 
% correctly include the many-body corrections through the scissors operator in 
% the SH susceptibility. Later, Cabellos \textit{et al}.\cite{cabellosPRB09} 
% elaborated a derivation of the ``velocity gauge'' or 
% ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the scissors operator 
% and proved gauge invariance with respect to the length gauge. From these works 
% it is clear the length gauge is a much better staring point to obtain the 
% surface SH (SSH) susceptibility, as will be elaborated in this article. 
% However, these considerations are only valid for bulk semiconductors.

% Concerning the optical response of surfaces and interfaces, Reining 
% \textit{et al.}\cite{reiningPRB94} introduced the concept of a cut function 
% in order to obtain the surface SH susceptibility tensor. This cut function 
% is required since one usually uses a slab approach when treating semi-infinite 
% surface systems. If the slab is centrosymmetric the susceptibility tensor will 
% be identically zero. The cut function is such that it separates the nonlinear 
% response for the two surfaces of the slab avoiding the destructive 
% interference between them giving a finite value that one identifies with the 
% SSH susceptibility tensor. If the slab is not centrosymmetric the cut function 
% can be used to separate the different signals coming from either surface of 
% the slab. Indeed, one of the main results of this article is to show that the 
% SSH  susceptibility tensor obtained by using the cut function is correctly 
% extracted from the slab. After Reining \textit{et al.},\cite{reiningPRB94}
% Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02,sanoPRB02}
% followed upon this work and in particular Ref.~\onlinecite{arzatePRB01} went 
% into a  detailed analysis of the different contributions to the SHG spectra 
% of a surface and the nuanced relationship between bulk, surface, interband, 
% intraband, $1\omega$ and $2\omega$ terms, and
% Ref. \onlinecite{mejiaRMF04} developed a layer-by-layer 
% analysis for the nonlinear responses of semiconductor systems, within a 
% tight-binding framework. This model allows for obtaining results from 
% selected regions of a system including the surface. However, in these 
% references the scissors operator is either excluded or incorrectly 
% implemented.

% Finally, DFT-LDA calculations are often based on the use of pseudopotentials.
% As we will discuss later in this article, the presence of the nonlocal part 
% of the pseudopotential introduces corrections to the momentum operator of 
% the electron that have to be included with care into the SSH susceptibility. 
% So far, this contribution has been neglected in the literature.

% Therefore, within the IPA the most complete approach for the calculation of 
% the SSH susceptibility is one which includes i) the scissors correction, 
% ii) the cut function, and iii) the contribution of the nonlocal part of the 
% pseudopotential. Therefore, the goal of this paper is to derive a new 
% expression within the length gauge for the SSH susceptibility tensor 
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ 
% that include the aforementioned contributions. The inclusion of these three 
% contributions makes our scheme unprecedented and opens the possibility to 
% study surface SHG with more versatility and providing more accurate results 
% than previous approaches.

% The paper is organized as follows. In Sec. \ref{theory} we present the 
% relevant steps for deriving the surface second-order susceptibility tensor
% $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
% length gauge formalism. This derivation includes the addition of the terms 
% mentioned above that have been absent in previous works. If desired, we can 
% use $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ for a 
% layer-by-layer analysis. In Sec. \ref{results} we show the results of this 
% reformulation with a study of a clean Si(001) surface with a $2 \times 1$ 
% reconstruction that proves the validity of our approach with a special 
% test-case that has never been implemented before. We compare results from 
% before and after adding the different nonlocal contributions. Finally, in 
% Sec. \ref{conc} we submit our conclusions.

@@

In this section we present the scheme used to calculate the surface 
second-order nonlinear response using the length gauge formalism. Our 
derivation includes the aforementioned terms not included before for 
this response.

\subsection{Perturbation approach}

We assume the IPA, a classical electromagnetic field, and quantum mechanical 
matter. We neglect local field and excitonic effects. We can describe the 
system using 
%a scaled 
the one electron density operator ${\rho}$, with which we can calculate the 
expectation value of a single-particle observable $\mathcal{O}$ as 
$\langle{\cal O}\rangle = \mbox{Tr}({\rho}{\cal O}),\label{calo}$
with $\mathcal{O}$ the associated quantum mechanical operator and Tr the 
trace. The density operator satisfies
$i\hbar(d{\rho}/dt) = [H(t), {\rho}], \label{rho}$
with $H(t)$ as the total single electron Hamiltonian, written as 
\begin{equation*}
H(t) = H_{0} + H_{I}(t),  
\label{ache}
\end{equation*}
where $H_{0}$ is the unperturbed time-independent Hamiltonian, and $H_{I}(t)$
is the time-dependent potential energy due to the interaction of the electron 
with the electromagnetic field.
% ; $H_{0}$ has eigenvalues
% $\hbar \omega_{n}(\mathbf{k})$
% and eigenstates $| n\mathbf{k} \rangle$ (Bloch states) labeled by a band 
% index $n$ and crystal momentum $\mathbf{k}$.
To proceed with the solution of $\rho$ it is convenient to use the interaction 
picture, where a unitary operator $U = \exp({iH_{0}t/\hbar})$ transforms any 
operator $\mathcal{O}$ into $\tilde{\mathcal{O}} = U{\cal O}U^{\dagger}$. 
Even if $\mathcal{O}$ is time-independent, $\tilde{\mathcal{O}}$ is 
time-dependent through the explicit time dependence of $U$. The dynamical 
equation for $\tilde{\rho}$ is given by
\begin{equation*}
i\hbar \frac{d\tilde{\rho}}{dt} = [\tilde{H}_{I}(t), \tilde{\rho}],
\label{rho1}
\end{equation*}
with solution 
\begin{equation}
i\hbar \tilde{\rho}(t) = 
    i\hbar \tilde{\rho}_{0} + 
    \int_{-\infty}^{t}dt^{\prime}
    [\tilde{H}_{I}(t^{\prime}),\tilde{\rho}(t^{\prime})],  
\label{trans}
\end{equation}
where $\tilde{\rho}_{0} = \tilde{\rho}(t = -\infty)$ is the unperturbed 
density matrix. We look for the standard perturbation series solution, 
$\tilde{\rho}(t) = \tilde{\rho}^{(0)} + \tilde{\rho}^{(1)} + 
\tilde{\rho}^{(2)} + \cdots$, where the superscript denotes the order (power) 
with which each term depends on the perturbation $H_{I}(t)$. From 
Eq.~\eqref{trans} the $N$-th order term is 
\begin{equation}
\tilde{\rho}^{(N)}(t) = 
    \frac{1}{i\hbar}\int_{-\infty }^{t}dt^{\prime}
    [\tilde{H}_{I}(t^{\prime }), \tilde{\rho}^{(N - 1)}(t^{\prime})].
\label{rhop}
\end{equation}
The series is generated by the unperturbed density operator $\tilde{\rho}^{(0)}
\equiv \tilde{\rho}_{0}$, assumed to be the diagonal Fermi-Dirac distribution, 
$\langle n\mathbf{k}|\tilde{\rho}_{0}|n\mathbf{k}\rangle = 
f(\hbar\omega_{n}(\mathbf{k}))\equiv f_{n}$. For a clean, cold semiconductor 
$f_{n}=1$ when $n$ is a valence ($v$) or occupied band, and zero when $n$ is 
a conduction ($c$) or empty band. We assume this for the remainder of the 
article. The expectation values must satisfy
$\langle{\cal O}\rangle = \mbox{Tr}({\rho}{\cal O}) = 
\mbox{Tr}(\tilde{\rho}\tilde{\cal O})$.

>>>>

\documentclass[floatfix,prb,aps,superscriptaddress,showpacs,11pt,preprint,letterpaper]{revtex4}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[backref,pdffitwindow,colorlinks,linkcolor={blue},citecolor={red}]{hyperref}
%\usepackage{showlabels}

@@

\begin{abstract}
We formulate a theoretical approach of surface second-harmonic generation 
from semiconductor surfaces based on the length gauge and the 
electron  
density operator.  
Within the independent particle approximation 
the nonlinear second-order surface susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$    
is calculated, including for the first time in one unique formulation
(i) the scissors correction, needed to have the correct value of the
energy band gap,
(ii) the contribution of the nonlocal part of the  
pseudopotentials, routinely  used in \textit{ab initio} band structure calculations, 
and (iii) the derivation for the inclusion of the cut function, 
used to extract the surface response.
The first two contributions
are described by spatially nonlocal quantum
mechanical operators and are fully taken into account in the present formulation. 
As a test-case of the approach we calculate 
$\chi^{xxx}(-2\omega;\omega,\omega)$ for the clean Si(001)$2\times 1$
reconstructed surface and find that our scheme to extract the surface
through the cut function
readily works and is exact. The effects of the scissors correction and
of the nonlocal part of the pseudopotentials are discussed 
in surface nonlinear optics. 
The scissors correction shifts the spectrum to 
higher energies though the shifting is not rigid and mixes the 
$1\omega$ and $2\omega$ resonances, 
and has a strong influence in the line-shape. 
The 
effects of the nonlocal part of the pseudopotentials 
keeps the same line-shape of $|\chi^{xxx}_{2\times 1}(-2\omega;\omega,\omega)|$, but 
reduces its value
by 15-20\%.  
Therefore, the inclusion of the three aforementioned 
contributions is very important and 
makes our scheme unprecedented and opens the possibility to 
study surface second-harmonic generation
with more versatility and providing more accurate results.
\end{abstract}

\maketitle

@@

In recent years surface nonlinear optical spectroscopies, particulary surface 
second-harmonic generation (SSHG), have evolved as useful nondestructive 
and noninvasive tools to study surface and interface properties. These 
properties include atomic structure, phase transitions, adsorption of 
atoms, and many others.\cite{daumPRL93, mcgilpOE94, meyerPRL95, powerPRL95, 
godefroyAPL96, hoferAPA96, dadapPRB97, bloembergenAPB99, mcgilpSRL99, 
suzukiAPB99, mitchellSS01, hughesPRB96, guyotPRB88, downerPSSA01, shenAPB99, 
shenNAT89, chenPRL81, mendozaPRL98, downerSIA01} Nowadays, SSHG spectroscopy 
is a crucial tool for research and development in microelectronics,
\cite{zheltikovLP00} semiconductors, \cite{lupkeSSR99} nanomaterials,
\cite{salazar-aparicioPRB14} and many more recent areas of great scientific 
and commercial interest.\cite{cazzanelliNM14} The high surface sensitivity 
of SSHG spectroscopy is due to the fact that within the dipole approximation 
the bulk SHG signal of centrosymmetric materials is identically zero.
Radiation can occur only at the surface where the inversion symmetry is 
broken. 
The bulk quadrupole contribution for centrosymmetric materials is
different from zero, but usually it is very small,\cite{downerSIA01} and we neglect it. 
Much of the foundation of surface science has been built from 
experiments involving emission or scattering of electrons from surfaces. 
These require ultrahigh vacuum (UHV) environments and provide no access to 
buried interfaces. However, SSHG is compatibile with non-UHV conditions and 
has access to interfaces buried beneath transparent overlayers. Even when 
applied to surfaces in UHV, the light source and detectors can be aligned and
used outside the vacuum chamber.

The usefulness of SSHG could be limited by the lack of microscopic theoretical
understanding of the nonlinear spectra. The macroscopic phenomenological 
theory of SSHG, which relates the intensity, phase, and polarization of 
detected fields to nonlinear and linear susceptibilities at the material 
interface is now fully developed.\cite{downerSIA01}
 However, microscopic theory that relates 
electronic-level structure to the nonlinear source polarization is still being 
developed. \cite{butcherPOPS63, aspnesPRB72, sipePRB93,
  levinePRB94,aversaPRB95, hughesPRB96, rashkeevPRB98,beyond}  
 Within the independent 
particle approximation (IPA) some frameworks for bulk SHG have been developed 
to study the nonlinear optical response of bulk materials. \cite{butcherPOPS63, aspnesPRB72, sipePRB93, levinePRB94,aversaPRB95, hughesPRB96, rashkeevPRB98} 
In this article we put forward an approach to calculate the
microscopic second-harmonic surface susceptibility that encompasses several
theoretical features not taken into account before 
in the case of a surface.

The most used framework for \textit{ab initio} calculations, Density 
Functional Theory (DFT) within the Local Density Approximation (LDA),
\cite{kohnPR65} underestimates the energy band gap of semiconductors. It is 
well understood that one has to include the many-body interaction to correct 
for this underestimation of the gap. In this context, the so-called GW 
approximation\cite{onidaRMP02} is known to correct the electronic gap of most 
semiconductors\cite{luceroJPCM12}. However this can be a very expensive
calculation and thus one uses the much simpler scissors operator scheme.
\cite{levinePRL89,levinePRL91,delsolePRB93}  
This allows us to ``open'' the DFT-LDA gap to 
its correct experimental or GW value even when experiments are not available. 
In this context, to correct for the underestimation of the energy band gap of 
semiconductors Nastos et al.\cite{nastosPRB05} used the ``length gauge'' or 
``$\mathbf{r}\cdot\mathbf{E}$ gauge'' to show how to correctly include the 
many-body corrections through the scissors operator in the SH susceptibility.
Later, Cabellos \textit{et al}.\cite{cabellosPRB09} elaborated a derivation 
of the ``velocity gauge'' or ``$\mathbf{A}\cdot\mathbf{v}$'' gauge properly including the 
scissors operator and proved gauge invariance with respect to the length 
gauge. From these works it is clear the length gauge is a much better starting
point to obtain the surface SH (SSH) susceptibility, as will be elaborated 
in this article. However, these considerations are only valid for bulk 
semiconductors.

Concerning the optical response of surfaces and interfaces, Reining 
\textit{et al.}\cite{reiningPRB94} introduced the concept of a cut function 
in order to obtain the surface SH susceptibility tensor. This cut function 
is required since one usually uses a slab approach when treating semi-infinite 
surface systems.\cite{reiningPRB94} If the slab is centrosymmetric the susceptibility tensor will 
be identically zero. The cut function is such that it separates the nonlinear 
response for the two surfaces of the slab avoiding the destructive 
interference between them giving a finite value that one identifies with the 
SSH susceptibility tensor. If the slab is not centrosymmetric the cut function 
can be used to separate the different signals coming from either surface of 
the slab. Indeed, one of the main results of this article is to show that the 
SSH  susceptibility tensor obtained by using the cut function is correctly 
extracted from the slab. After Reining \textit{et al.},\cite{reiningPRB94}
Refs.~\onlinecite{mendozaPRL98,arzatePRB01,mendozaPRB01,mejiaPRB02,sanoPRB02}
followed 
upon this work and in particular Ref.~\onlinecite{arzatePRB01} went into a 
detailed analysis of the different contributions to the SHG spectra of a 
surface and the nuanced relationship between bulk, surface, interband, 
intraband, $1\omega$ and $2\omega$ terms, and
Ref. \onlinecite{mejiaRMF04} developed a layer-by-layer 
analysis for the nonlinear responses of semiconductor systems, within a 
tight-binding framework. This model allows for obtaining results from 
selected regions of a system including the surface. However, in these 
references the scissors operator is either excluded or incorrectly 
implemented.

Finally, DFT-LDA calculations are often based on the use of pseudopotentials.
As it will be discussed 
in this article, the presence of a nonlocal part of the
pseudopotential introduces
corrections to
the momentum operator of the electron that have to be included with care in 
the SSH susceptibility. 
For the bulk counterpart see for instance Refs.~\onlinecite{ismailPRL01,luppiPRB08}.

Therefore, within the IPA the most complete approach for the calculation of 
the SSH susceptibility is one which includes (i) the scissors correction, 
(ii) the contribution of the nonlocal part of the 
pseudopotential, and
(iii) the cut function.
 Therefore, the goal of this paper is to derive a new 
expression within the length gauge for the SSH susceptibility tensor 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ 
that includes the aforementioned contributions. The inclusion of these three 
contributions makes our scheme unprecedented and opens the possibility to 
study surface SHG with more versatility and providing accurate results.

The paper is organized as follows. In Sec. \ref{theory} we present the 
relevant steps for deriving the surface second-order susceptibility tensor
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ within the 
length gauge formalism. This derivation includes the addition of the terms 
mentioned above that have been absent in previous works. Also, our 
$\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$ can be used 
for a layer-by-layer analysis if desired. 
Nevertheless, such an analysis could depend on the choice of the basis
used to expand the wavefunctions.
In Sec. \ref{results} we show the 
results of this reformulation with a study of a clean Si(001) surface with a 
$2 \times 1$ reconstruction, 
proving the correctness of our approach with a special test-case,
never exploited before.
We compare results from before and after adding 
the different nonlocal contributions. Finally, in Sec. \ref{conc} we give 
our conclusions.

@@

In this section we present the scheme used to calculate the surface 
second-order nonlinear response using the length gauge formalism and
the electron 
density operator. 
Some of the results presented  in Subsections \ref{pa} and \ref{lgf}
have already been discussed in earlier
studies and can be applied to bulk and surface studies.\cite{sipePRB00,cabellosPRB09}
We present them in order to have a self-contained derivation.
The terms presented in these Subsections are used in Subsection \ref{cd}, 
where we derive the expression 
for the surface $\chi^{\mathrm{a}\mathrm{b}\mathrm{c}}(-2\omega;\omega,\omega)$.

\subsection{Perturbation approach}\label{pa}
